Dependent Object Types

### 2012-09-22 doc/fool/article.tex
type safety intuition

We propose DOT as a core calculus for path-dependent types. We present
the calculus formally in section \ref{dot-calculus} and through
examples in section~\ref{dot-examples}. Though we show that the
calculus does not satisfy the standard theorem of preservation in
section~\ref{dot-preservation}, we contribute a proof sketch of type
safety using logical relations in section~\ref{dot-type-safety}. In
section~\ref{discussion}, we discuss choices and variants of the
calculus, as well as related work, and conclude in
section~\ref{conclusion}.

@@

We sketch a proof that the DOT calculus is type-safe using logical
relations in section~\ref{dot-type-safety}. However, we first tried to
prove the calculus type-safe using the standard theorems of
preservation and progress~\cite{soundness,tapl}. Unfortunately, for
the calculus as presented, and any of its variants that we devised,
preservation doesn't hold. In this section, we review some of the most
salient counterexamples to preservation that we found. These
counterexamples have been checked with PLT Redex~\cite{plt_redex}.

@@

>>>>

We propose DOT as a core calculus for path-dependent types. We present
the calculus formally in section \ref{dot-calculus} and through
examples in section~\ref{dot-examples}. In
section~\ref{dot-preservation}, we show that the calculus does not
satisfy the standard theorem of preservation, also known as subject
reduction. However, we still believe that the calculus is type-safe,
as discussed in section~\ref{dot-type-safety}. In
section~\ref{discussion}, we discuss choices and variants of the
calculus, as well as related work, and conclude in
section~\ref{conclusion}.

@@

We first tried to prove the calculus type-safe using the standard
theorems of preservation and
progress~\cite{soundness,tapl}. Unfortunately, for the calculus as
presented, and any of its variants that we devised, preservation
doesn't hold. In this section, we review some of the most salient
counterexamples to preservation that we found. These counterexamples
have been checked with PLT Redex~\cite{plt_redex}.

@@

\section{Type Safety via Logical Relations}\label{dot-type-safety}

We believe that the DOT calculus is type-safe, and are developing a
proof of type safety using step-indexed logical
relations~\cite{ahmed04,ahmed06,step_indexed_obj}. In this section, we
briefly summarize the main theorem of type safety, and the strategy
based on step-indexed logical relations that we are using to prove
it. All our development, including models in Coq and PLT Redex, is
available from \url{http://dot.namin.net}.

\subsection{Type Safety}
Type safety states that a well-typed program doesn't get stuck. More
formally: If $\emptyset \ts t \typ T$ and $\reductionl t \emptyset
{t'} {s'} *$ then either $t'$ is a value or $\exists t'',
s''. \reduction {t'} {s'} {t''} {s''}$. Note that this is stronger
than the standard theorem of progress, which states that a well-type
term can take \emph{a} step or is a value.

\subsection{Strategy based on Logical Relations}
Our strategy follows the standard technique of proving type safety
using logical relations. We define a logical relation $\Gamma \ds t
: T$, such that $\Gamma \ts t : T$ implies $\Gamma \ds t : T$ implies
type safety.

The main logical relation $\Gamma \ds t : T$ is based on a set of
mutually recursive logical relations that are step-indexed in order to
ensure their well-foundedness: $\rele k \Gamma s T$ defines the set of
terms that appear to have type $T$ when taking at most $k$ steps, and
$\relv k \Gamma s T$ defines the set of \emph{values} that appear to
have type $T$ when taking at most $k$ steps. There are also two other
logical relations, one for completing the store to match the context,
and one for completing the context to match the store resulting from
taking some number of steps.

$\Gamma \ds t : T$ is defined as $t \in \rele k \Gamma \emptyset T,
\forall k$. $t \in \rele k \Gamma s T$ roughly as follows: after
reducing $t$ a number of steps $<k$, if the resulting term is
irreducible, then it must be in $\relv {j'} {\Gamma'} {s'} T$ for
appropriate $j', \Gamma',s'$. By definition, $\Gamma \ds t : T$
implies that $t$ cannot be stuck.

Then, to prove type safety, all that needs to be proved is the
fundamental theorem -- or completeness -- of the logical relation:
$\Gamma \ts t : T$ to $\Gamma \ds t : T$. Type safety is a
straightforward corollary of this theorem, since $\Gamma \ds t : T$
implies by definition that $t$ cannot be stuck. The proof of the
fundamental theorem is on induction on the derivation of $\Gamma \ts t
: T$. The logical relation approach enables us to have strong enough
induction hypotheses to carry the proof through, without requiring us
to strictly relate intermediate terms by types like preservation.
